---
title: "Distances and Multidimensional Scaling (MDS)"
author: "Masha"
date: "2/6/2020"
output: 
   md_document:
    variant: markdown_github
---

```{r}
library (MASS)
#library(rgl)
```

#Dataset
```{r}
DACA = read.csv("C:/Users/fb8502oa/Desktop/DSCI 415/DSCI 415 fall/DACA.csv", header=TRUE)
head(DACA)
```

1.Use the raw data, which is mostly counts, to create metrics

#select variables v1, v2, v3 and subsetting 
#I am interested to see how the lawsuits were approved depending per race. Eg:
  #1.Approved per white vs Approved per black.
  #2.Approved per Hispanic and Approved per Asian.

```{r}
my_data=DACA[,c(17:20)]
```


2. Using the data and the metrics I created to conduct a metric MDS (or non-metric MDS) with k=2 dimensions, K=3 dimensions.   
 Finding the stress and Constructing a plot of the dimensions created and label the points by state, coloring the points by lawsuit status. 

```{r}
##scaling the data is very important
Daca.scale =scale(my_data) 
##now finding the distance 
Dacadis = dist(Daca.scale)
##k=2
Daca.samm = sammon(Dacadis,k=2)
```

```{r echo=TRUE, fig.height=10, fig.width=10}
#plotting
plot(Daca.samm$points,type="n")
text(Daca.samm$points,as.character(DACA$State),col=as.numeric(DACA$Lawsuit)+2) ##the 2 is for the coloring. 
##The very blue ones are lawsuits against obama, the light blue are against trump while the green is no lawsuits.
```
comments for #2. 
The outliers from
sammon make sense since California was among the leading state in lawsuits against trump
while Texas had lawsuits against Obama.
The dark blue are the states that have lawsuits against Obama while the light blue states
have lawsuits against trump while the green states had none.
The initial stress is key = 0.03017


3.Repeat part (2) using 𝑘 = 3 dimensions

```{r eval=FALSE, include=FALSE}
##k=3
Daca.samm3d = sammon(Dacadis,k=3) 
##install plot 3-d (rgl)
plot3d(Daca.samm3d$points,type="n")
text3d(Daca.samm3d$points, texts= as.character(DACA$State),col=as.numeric(DACA$Lawsuit)+3)
```


Problem 2 - Gene Expression Levels for Colon Tissue Samples
These data come from a microarray experiment where gene expression levels were measured
for colon tissues samples from 40 individuals with colon cancer and 22 individuals without
colon cancer. Below is a blurb about microarray experiments taken from Modern Multivariate
Statistical Techniques by Alan Izenman.

#The dataset
```{r}
Alontop = read.csv("C:/Users/fb8502oa/Desktop/DSCI 415/DSCI 415 fall/Alontop.csv")
View(Alontop)
```

Form a distance matrix for the 62 tissue samples using Euclidean distance. Is scaling
important? Why or why not? Give summary statistics for the pairwise distances in the
distance matrix.
```{r}
names(Alontop)
alon.mat = Alontop[,-93]

dist2 = as.matrix(dist(alon.mat))
dist2

summary(as.numeric(dist2))
```

Using the distance matrix 𝐷 formed in part (a) conduct a metric (classical) multidimensional
scaling using 𝑘 = 2 dimensions. What is the stress of this representation? Construct a plot
of the two dimensions found using metric MDS and label + color the points in the plot by
Tissue Type. Discuss what you learn from this plot
```{r}
dist2.samm= isoMDS(dist2, k=2)

plot(dist2.samm$points, type="n")
text(dist2.samm$points, labels = as.character(Alontop$Tissue.Type), col= as.numeric(Alontop$Tissue.Type)+2, cex=0.6)
title(main = "metrics for cell cancers (k=2)")

```

Repeat all aspects of part (b) using 𝑘 = 3 dimensions. Does the third dimension help
differentiate the cancerous and benign tumor tissue samples?
```{r}

#for the 3d
summary(as.numeric(dist2))
dist2.samm3= isoMDS(dist2, k=3)
```

```{r eval=FALSE, include=FALSE}
plot3d(dist2.samm3$points, type="n")
text3d(dist2.samm3$points, texts = as.character(Alontop$Tissue.Type), col= as.numeric(Alontop$Tissue.Type)+2, cex=0.6)
#title(main = "metrics for cell cancers (k=3)")
```


comment = 
When using sammon, the stress levels are lower but with a higher dimension, the stress levels
also increase. In this case, the 2d representation is way better than the 3d. However, when using
isoMDS scaling, the stress on the 3d representation is was smaller compared to the 2d
representation


d.) Using the sample correlation matrix (𝑅) as a measure of variable similarity form a distance
matrix from it (use one of the schemes for turning similarity measures into a dissimilarity
measures). Using this variable distance matrix conduct a non-metric MDS using 𝑘 = 2
dimensions. Construct a plot of these dimensions labeling the points using the gene names.
Do there are appear to groups of genes that express together? If so, pick one of the groups
a small number of members and list the genes that make up that group


```{r}
library(corrplot)
alon.cor = cor(alon.mat)
#alon.cor
corrplot(alon.cor)
```

```{r}
#cor for 2d
var.dist= sqrt(2*(1-alon.cor))
var.mds=cmdscale(var.dist, k=2)
```





```{r}
plot(var.mds,xlab = "dim 1",ylab ="dim 2", type ="n")
text(var.mds,labels = names(alon.mat))
```


comment 
genes in the same group.
• M80815
• D29808
• H64489
• Z49269.2
• Z49269
• F71025
• U19969
• M76378.2
• M7678
• M7678.3
• M63391

e.)
Repeat all aspects of part (d) using 𝑘 = 3 dimensions.
```{r}
#cor for 3d
var.dist3= sqrt(2*(1-alon.cor))
var.mds3=cmdscale(var.dist, k=3)
```



```{r eval=FALSE, include=FALSE}
plot3d(var.mds3,xlab = "dim 1",ylab ="dim 2", type ="n")
text3d(var.mds3,texts = names(alon.mat))
```





Problem 3 – Dune Meadow Species Communities
In their paper “Data Analysis in Community and Landscape Ecology” Jongman, R.H.G, ter Braak,
C.J.F & van Tongeren, O.F.R., present the results from a study examining 20 dune meadow sites
and counting the presence of 30 different species found within these sites. There were also
environmental features of the sites measured such as moisture, soil depth, and land
management. These data are contained in the file: Dune with Environmentals.csv.
Using the Bray-Curtis measure throughout, form distance matrices for site-to-site (row)
differences and species-to-species (variables) differences, then carry out a thorough MDS of
these data for sites and species. Create plots of your MDS results labeling sites, or species,
whichever is appropriate. Use the environmental factors to add more information to these
displays through use of color-coding or bubbles (or both).

 for the site plot(s) and 10 pts. for the species MDS plot(s)
Variables:
• Site – site number 1 – 20
• A1 – thickness of soil A1 horizon
• Moisture – ordered factor denoting moisture level 1 < 2 < 4 < 5
• Management - a factor with levels: BF (Biological farming), HF (Hobby farming), NM
(Nature Conservation Management), and SF (Standard Farming)
• Use – ordered factor of land use Hayfield < Haypastu < Pasture
• Manure – ordered factors with levels 0 < 1 < 2 < 3 < 4\
• Achimill – Callcusp – species counts at each site

```{r}
Dune <- read.csv("C:/Users/fb8502oa/Desktop/DSCI 415/DSCI 415 fall/Dune with Environmentals.csv")
View(Dune)
```

```{r}
#dim(Dune)
spec.mat= Dune[,-c(1:6)]
head(spec.mat)
spec4.mat= spec.mat[, colSums(spec.mat)>4]
colSums(spec4.mat)
rowSums(spec4.mat)
dim(spec4.mat)
```

```{r}

library(proxy)
site.dist= dist(spec4.mat,method="Bray")
site.iso = isoMDS(site.dist, k=2)
plot(site.iso$point, type="n")
text(site.iso$point, as.character(Dune$Site), col= as.numeric(Dune$Management)+1)
```

```{r}

#transposing the matrix
tspec4.mat= t(spec4.mat)
spec.dist =dist(tspec4.mat, method= "Bray")
spec.iso = isoMDS(spec.dist, k=2)
plot(spec.iso$points, type="n")
text(spec.iso$points, as.character(row.names(tspec4.mat)),cex=0.6)
```




















